---
title: "thesis_William_steps"
author: "chiara francesca ausenda"
date: "2025-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Library load
```{r}
library(tidyverse)
library(sf)
library(RNetCDF)
library(dplyr)
library(lubridate)
library(ggplot2)
library(rmapshaper)
```


## Step 1 - Get EEZ polygons
okay so basically the problem is with the high seas area. I need to plot both the bounding box and the eez thing to see if ACTUALLY there is an high seas area at all or it's all good-> tried lines 60-65 but i don't think it's the correct code-> check that and then try to solve.
```{r}
# Read EEZ Geopackage file
eez_sf <- st_read("~/GitHub/Ausenda-Wernick/World_EEZ_v12_20231025_LR/eez_v12_lowres.shp")

# Ensure no use of spherical geometry for validity checks
sf_use_s2(FALSE)

# Validity check for EEZ polygons
#valid_check <- st_is_valid(eez_sf)
#all(valid_check)

# NOT NEEDED - Fix any invalid geometries in EEZ
#eez_sf <- st_make_valid(eez_sf)

# Drop the specified columns
eez_sf <- eez_sf %>%
  select(-c(TERRITORY1, MRGID_TER1, UN_TER1, MRGID_SOV1, UN_SOV1, 
            AREA_KM2, MRGID_TER2, MRGID_TER3, MRGID_SOV2, MRGID_SOV3, 
            UN_TER2, UN_TER3, UN_SOV2, UN_SOV3))

# Define bounding box based on new coordinates for EEZ
lon_min <- -95.3388  # westernmost longitude (Ecuador - Galapagos Islands EEZ)
lon_max <- -25.294   # easternmost longitude (Brazil - Trindade Islands EEZ)
lat_min <- -59.8527  # southernmost latitude (Chile EEZ)
lat_max <- 16.7469   # northernmost latitude (Venezuela EEZ)

# Create bounding box polygon
bbox_polygon <- st_sfc(st_polygon(list(matrix(c(lon_min, lat_min,
                                                 lon_min, lat_max,
                                                 lon_max, lat_max,
                                                 lon_max, lat_min,
                                                 lon_min, lat_min),
                                               ncol = 2, byrow = TRUE))), crs = st_crs(eez_sf))

#Perform intersection: retain only EEZ polygons within the bounding box (intersecting with the box)
eez_sf_bbox <- st_intersection(eez_sf, bbox_polygon)

#visualization:
#ggplot(eez_sf_bbox) + geom_sf()

```
If I want high seas, but there's also land lol

high_seas <- st_difference(bbox_polygon, st_union(eez_sf_bbox))
ggplot(high_seas) + geom_sf()




##Temperature Anomalies and data manipulation

```{r}
# Open dataset
url_V5 <- 'https://www.ncei.noaa.gov/thredds/dodsC/noaa-global-temp-v5/NOAAGlobalTemp_v5.0.0_gridded_s188001_e202212_c20230108T133308.nc'
data_V5 <- open.nc(url_V5)

# Extract variables
latitude <- var.get.nc(data_V5, 'lat')
longitude <- var.get.nc(data_V5, 'lon')

# Convert longitude from 0-360 to -180 to 180 if needed
longitude_shifted <- ifelse(longitude > 180, longitude - 360, longitude)

# Convert time to Date format (days since 1800-01-01)
time <- var.get.nc(data_V5, "time")
time_dates <- as.Date("1800-01-01") + time
time_df <- data.frame(time_index = 1:length(time), date = time_dates)

# Filter time range from 1990 to 2022
time_df <- time_df %>% filter(year(date) >= 1990 & year(date) <= 2022)

# Group by approx months (30-day bins) and extract anomalies
monthly_avg <- time_df %>%
  mutate(month_year = format(date, "%m-%Y")) %>%
  group_by(month_year) %>%
  summarise(time_index = list(time_index), .groups = 'drop')

# Function to create a polygon from a central point (lon, lat)
create_polygon <- function(lon, lat, res = 5) {
  coords <- matrix(
    c(lon - res/2, lat - res/2,  # Bottom-left
      lon + res/2, lat - res/2,  # Bottom-right
      lon + res/2, lat + res/2,  # Top-right
      lon - res/2, lat + res/2,  # Top-left
      lon - res/2, lat - res/2), # Closing the polygon
    ncol = 2, byrow = TRUE
  )
  return(st_polygon(list(coords)))
}

# Initialize list to store results
results_list <- list()

# Loop through each month-year group to extract and average anomaly values
for (i in seq_len(nrow(monthly_avg))) {
  indices <- unlist(monthly_avg$time_index[i])
  anom_values <- sapply(indices, function(idx) var.get.nc(data_V5, 'anom', start=c(NA, NA, 1, idx), count=c(NA,NA,1,1)))
  
  # Compute average anomaly for the month
  avg_anom <- rowMeans(anom_values, na.rm = TRUE)
  
  # Create spatial dataframe
  df <- expand.grid(lon = longitude_shifted, lat = latitude)
  df$anom <- as.vector(avg_anom)
  df$date <- monthly_avg$month_year[i]
  df$geometry <- mapply(create_polygon, df$lon, df$lat, SIMPLIFY = FALSE)
  
  # Subset the dataframe based on bounding box
  df <- df %>% filter(lon >= lon_min & lon <= lon_max & lat >= lat_min & lat <= lat_max)
  
  # Store in list
  results_list[[i]] <- df
}

# Combine all results into a single sf object
t_anom_sf <- do.call(rbind, results_list) %>% st_as_sf(crs = 4326)

# Print first few rows
#print(head(t_anom_sf))

```

##Spatial join t_anom_sf in eez_sf_bbox
```{r}
# Perform spatial join: Assign anomalies from t_anom_sf to EEZ polygons
sf_use_s2(FALSE)
eez_t_anom <- st_join(t_anom_sf, eez_sf_bbox, left = TRUE)  # Keep all EEZ polygons

# Group by EEZ and date, then compute the average anomaly
eez_anom <- eez_t_anom %>%
  group_by(MRGID_EEZ, date) %>%  # MRGID_EEZ uniquely identifies EEZs
  summarise(avg_anom = mean(anom, na.rm = TRUE), .groups = 'drop')

# Merge anomaly data back into EEZ polygons
eez_t_final <- st_join(eez_sf, eez_anom, by = "MRGID_EEZ")

# Print first few rows
print(head(eez_t_final))
```

#Formatting eez_t_final
```{r}
selected_countries <- c("VEN", "COL", "ECU", "PER", "CHL", "ARG", 
                        "URY", "BRA", "GUY", "SUR", "FRA")

eez_t_final <- eez_t_final %>% 
  filter(ISO_SOV1 %in% selected_countries) %>% 
  select(c(ISO_SOV1, date, avg_anom)) %>% 
  rename(eez_country = ISO_SOV1) %>% 
  mutate(date = format(dmy(paste("01-", date, sep="")), "%Y-%m"))  # Convert MM-YYYY to YYYY-MM

eez_t_final <- st_drop_geometry(eez_t_final)

eez_t_final <- eez_t_final %>%
  group_by(date, eez_country) %>%
  summarize(avg_anom = mean(avg_anom, na.rm = TRUE), .groups = "drop")

print(head(eez_t_final))
```


##Transorfming and formatting piracy_df
```{r}
piracy_df <- read.csv("~/GitHub/Ausenda-Wernick/pirate_attacks.csv",
                      header = TRUE, stringsAsFactors = FALSE)

# Select countries
selected_countries <- c("VEN", "COL", "ECU", "PER", "CHL", "ARG", 
                        "URY", "BRA", "GUY", "SUR", "FRA")
piracy_df <- piracy_df %>% 
  filter(eez_country %in% selected_countries)

# Convert date variable to Date format
piracy_df$date <- as.Date(piracy_df$date, format = "%Y-%m-%d")

# Create year-month column
piracy_df <- piracy_df %>%
  mutate(date = format(date, "%Y-%m"))

complete_grid <- expand_grid(
  date = unique(piracy_df$date),
  eez_country = unique(piracy_df$eez_country)
)

piracy_count <- piracy_df %>%
  group_by(date, eez_country) %>%
  summarize(EventsCount = n(), .groups = "drop")

piracy_count <- complete_grid %>%
  left_join(piracy_count, by = c("date", "eez_country")) %>%
  mutate(EventsCount = replace_na(EventsCount, 0))

print(head(piracy_count))
```
##visualization

# Convert piracy data into spatial object using longitude & latitude
piracy_sf <- st_as_sf(piracy_df, coords = c("longitude", "latitude"), crs = 4326)

# Ensure both layers have the same CRS
piracy_sf <- st_transform(piracy_sf, crs = st_crs(t_anom_sf))


# Summarize data by counting occurrences of each year
piracy_summary <- piracy_df %>%
  mutate(year = year(as.Date(paste0(year_month, "-01")))) %>%  # Extract year from year_month
  count(year)  # Creates a new column 'n' with counts

# Plot with years on x-axis
ggplot(piracy_summary, aes(x = as.factor(year), y = n)) +
  geom_bar(stat = "identity", fill = "blue", color = "black") +
  labs(title = "Distribution of Piracy Events by Year", x = "Year", y = "Count") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Joining-> count_ym_anom
```{r} 
#join
count_ym_anom <- full_join(piracy_count, eez_t_final, by = c("date", "eez_country")) %>% 
  mutate(EventsCount = replace_na(EventsCount, 0))

# Print result
print(head(count_ym_anom))

write.csv(count_ym_anom, "count_ym_anom.csv", row.names = FALSE)
```


```{r}
count_ym_anom <- read.csv("~/GitHub/Ausenda-Wernick/count_ym_anom.csv")

# Plot the data
ggplot(count_ym_anom, aes(x = date, y = EventsCount)) +
  geom_line() +  # This will create a line plot
  geom_point() + # This will add points for each data point
  labs(title = "Events Count over Time",
       x = "Date",
       y = "Events Count") +
  theme_minimal()  # Optional: for a cleaner look

#look at monthly trends
count_ym_anom$date <- as.Date(paste0(count_ym_anom$date, "-01"))
count_ym_anom$Month <- factor(month(count_ym_anom$date, label = TRUE), levels = month.abb)

monthly_trend <- count_ym_anom %>%
  group_by(Month) %>%
  summarise(TotalEvents = sum(EventsCount, na.rm = TRUE))

ggplot(monthly_trend, aes(x = Month, y = TotalEvents, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 2) +
  labs(title = "Monthly Event Count Trend",
       x = "Month",
       y = "Total Events") +
  theme_minimal()


#look at yearly trends
count_ym_anom$year <- factor(year(count_ym_anom$date))

yearly_trend <- count_ym_anom %>%
  group_by(year) %>%
  summarise(TotalEvents = sum(EventsCount, na.rm = TRUE))

ggplot(yearly_trend, aes(x = year, y = TotalEvents, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 2) +
  labs(title = "Yearly Event Count Trend",
       x = "Year",
       y = "Total Events") +
  theme_minimal()
```

