---
title: "alt_route_IV"
author: "chiara francesca ausenda"
date: "2025-04-07"
output: html_document
---

```{r, include = F}
knitr::opts_chunk$set(results = 'markup')
```

#Load libraries
```{r}
library(tidyverse)
library(broom) 
library(modelsummary) 
library(knitr) 
library(kableExtra)
library(janitor) 
library(estimatr)
library(lubridate)
library(zoo)
library(pscl)  
library(texreg)
library(broom)
library(kableExtra)
library(sandwich)
```

#load data
```{r}
final_df <- read.csv("~/GitHub/Ausenda-Wernick/final_df.csv")

final_df$date <- as.Date(paste0(final_df$date, "-01"))

#F-statistics

final_df <- final_df %>% 
  filter(date >= as.Date("2017-01-01") & date <= as.Date("2019-12-31")) %>% 
    mutate(IUU_EventsCount = ifelse(is.na(IUU_EventsCount), 0, IUU_EventsCount)) %>% 
  mutate(
    EventsCount_lag = lag(EventsCount, 6),
  )
  
```

#2SLS
```{r, echo = TRUE, results = 'markup'}
summary(lm(IUU_EventsCount ~ avg_anom, data = final_df))

#iv model
late_model <- estimatr::iv_robust(EventsCount ~ IUU_EventsCount | avg_anom, data = final_df)


modelsummary::modelsummary(
  list("normal_IV" = late_model),
  stars = c('*' = .1, '**' = .05, '***' = .01),
  statistic = 'conf.int',
  coef_rename = c("(Intercept)" = "Constant", "net_use" = "Net use"),
  fmt = 2,
  model_titles = c("classic iv"),
  )
```

```{r eval=FALSE, include=FALSE}
#significant-> decent intrument even if low f stat
linear1 <- lm(IUU_EventsCount ~ avg_anom, data=final_df)

modelsummary::modelsummary(list("H2 linear model"=linear1),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2)




#no avg anom significance-> exogeneity of instrument
linear2 <- lm(EventsCount ~ avg_anom + IUU_EventsCount, data=final_df)

modelsummary::modelsummary(list("IV Checks"=linear2),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2)



```

#zero-inflated model
```{r}

# Create a clean dataset with complete cases
df_clean <- final_df %>% 
  na.omit()

# First-stage regression
first_stage <- lm(IUU_EventsCount ~ avg_anom , data = df_clean)

# Get predicted IUU and residuals
df_clean$pred_IUU <- predict(first_stage)
df_clean$resid_IUU <- resid(first_stage)

# Now run the zero-inflated Poisson model
zip_iv_model <- zeroinfl(EventsCount ~ pred_IUU + resid_IUU 
                         | avg_anom ,
                         data = df_clean,
                         dist = "poisson")
```

#logged version
```{r}
final_df$log_EventsCount <- log(final_df$EventsCount+1)
final_df$log_IUU_EventsCount <- log(final_df$IUU_EventsCount+1)

log_model <- estimatr::iv_robust(log_EventsCount ~ log_IUU_EventsCount | avg_anom, data = final_df)


modelsummary::modelsummary(list("log_IV"=log_model),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2)
```

#aggregated model
```{r}

#QUARTER
# Add a quarter column
final_df <- final_df %>%
  mutate(quarter = paste0(year(date), " Q", quarter(date)))

# Aggregate by eez_country and quarter
quarterly_df <- final_df %>%
  group_by(eez_country, quarter) %>%
  summarise(
    EventsCount = sum(EventsCount, na.rm = TRUE),
    IUU_EventsCount = sum(IUU_EventsCount),
    avg_anom = mean(avg_anom, na.rm = TRUE),
    .groups = "drop"
  )

agg_late_model <- estimatr::iv_robust(EventsCount ~ IUU_EventsCount | avg_anom, data = quarterly_df)


modelsummary::modelsummary(list("agg_IV"=agg_late_model),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2,
                           )
```


#YEAR-> IUU_eventscount intercept doesn't change so i dunno
# Extract year
final_df <- final_df %>%
  mutate(year = year(date))

# Aggregate by eez_country and year
yearly_df <- final_df %>%
  group_by(eez_country, year) %>%
  summarise(
    EventsCount = sum(EventsCount, na.rm = TRUE),
    IUU_EventsCount = sum(IUU_EventsCount, na.rm = TRUE),
    avg_anom = mean(avg_anom, na.rm = TRUE),
    .groups = "drop"
  )

y_agg_late_model <- estimatr::iv_robust(EventsCount ~ IUU_EventsCount | avg_anom, data = yearly_df)


modelsummary::modelsummary(list("LATE"=y_agg_late_model),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2)


#lagged models
```{r}
## 1-MONTH LAG
monthly_lagged_df<- final_df %>%
  arrange(eez_country, date) %>%  # Ensure the data is sorted by country and date
  mutate(
    EventsCount_lag1 = lag(EventsCount, 2),  # Lagged dependent variable
    IUU_EventsCount_lag1 = lag(IUU_EventsCount, 2)  # Lagged independent variable
  )
```

```{r}
#Normal IV

monthly_lag_iv <- iv_robust(EventsCount_lag1 ~ IUU_EventsCount_lag1 | avg_anom, data = monthly_lagged_df)

modelsummary::modelsummary(list("lag_IV"=monthly_lag_iv),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2)

```

```{r}
#lagged zero-inflated model

# Create a clean dataset with complete cases
lag_df_clean <- monthly_lagged_df %>% 
  na.omit()

# First-stage regression
first_stage <- lm(IUU_EventsCount_lag1 ~ avg_anom , data = lag_df_clean)

# Get predicted IUU and residuals
lag_df_clean$pred_IUU <- predict(first_stage)
lag_df_clean$resid_IUU <- resid(first_stage)

# Now run the zero-inflated Poisson model
lag_zip_iv_model <- zeroinfl(EventsCount_lag1 ~ pred_IUU + resid_IUU 
                         | avg_anom ,
                         data = lag_df_clean,
                         dist = "poisson")
```

```{r}
#lagged quarterly-aggregated iv
# Lag the EventsCount and IUU_EventsCount by 1 quarter
lag_quarterly_df <- quarterly_df %>%
  arrange(eez_country, quarter) %>%  # Ensure data is sorted by country and quarter
  mutate(
    EventsCount_lag1 = lag(EventsCount, 1),  # Lag EventsCount by 1 quarter
    IUU_EventsCount_lag1 = lag(IUU_EventsCount, 1)  # Lag IUU_EventsCount by 1 quarter
  )

agg_lag_late_model <- estimatr::iv_robust(EventsCount_lag1 ~ IUU_EventsCount_lag1 | avg_anom, data = lag_quarterly_df)


modelsummary::modelsummary(list("lag_agg"=agg_lag_late_model),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2)

```

#lag-log piracy
```{r}
monthly_lagged_log_df<- final_df %>%
  arrange(eez_country, date) %>%  # Ensure the data is sorted by country and date
  mutate(
    log_EventsCount_lag1 = lag(log_EventsCount, 2),  # Lagged dependent variable
    log_IUU_EventsCount_lag1 = lag(log_IUU_EventsCount, 2)  # Lagged independent variable
  )

monthly_lag_log_iv <- iv_robust(log_EventsCount_lag1 ~ log_IUU_EventsCount_lag1 | avg_anom, data = monthly_lagged_log_df)

modelsummary::modelsummary(list("lag_log_IV"=monthly_lag_log_iv),
                           stars = c('*' = .1, '**' = .05, '***' = .01),
                           statistic = 'conf.int',
                           coef_rename = c("(Intercept)"="Constant",
                                           "net_use" = "Net use"),
                           fmt = 2)
```


```{r}
#Colombia
colombia_df <- final_df %>% 
  filter(eez_country == "COL")

#iv model
COL_LATE <- estimatr::iv_robust(EventsCount ~ IUU_EventsCount | avg_anom, data = colombia_df)


modelsummary::modelsummary(
  list("COL_normal_IV" = COL_LATE),
  stars = c('*' = .1, '**' = .05, '***' = .01),
  statistic = 'conf.int',
  coef_rename = c("(Intercept)" = "Constant", "net_use" = "Net use"),
  fmt = 2,
  model_titles = c("classic iv")
  )

```

#look at distribution
colombia_df <- colombia_df %>%
  mutate(year = year(date)) %>% 

# Aggregate by eez_country and year
yearly_col_df <- colombia_df %>%
  group_by(year) %>%
  summarise(
    EventsCount = sum(EventsCount, na.rm = TRUE),
    IUU_EventsCount = sum(IUU_EventsCount, na.rm = TRUE),
    avg_anom = mean(avg_anom, na.rm = TRUE),
    .groups = "drop"
  )

print(ggplot(yearly_col_df, aes(x = factor(year), y = EventsCount)) +
     geom_col(fill = "darkorange") +
     labs(
         title = "Yearly Distribution of EventsCount in Colombia EEZ",
         x = "Year",
         y = "Total EventsCount") +
     theme_minimal())+
     theme(axis.text.x = element_text(angle = 45, hjust = 1))

#plots
```{r eval=FALSE, include=FALSE}
model_list <- list(late_model, log_model, agg_late_model, monthly_lag_iv, monthly_lag_log_iv, agg_lag_late_model)

htmlreg(model_list, file = "model_comparison3.html", 
         custom.coef.names = c("Intercept", "IUU_EventsCount", "log_IUU_EventsCount", "IUU_EventsCount_lag1", "log_IUU_EventsCount_lag1"),
         stars = 1,
         title = "Comparison of IV Models")



#zip

zip_coefficients <- summary(zip_iv_model)$coefficients
count_coefficients <- zip_coefficients$count
zero_coefficients <- zip_coefficients$zero

poisson_table <- data.frame(
  Predictor = rownames(count_coefficients),
  Estimate = count_coefficients[, "Estimate"],
  Std_Error = count_coefficients[, "Std. Error"],
  Z_Value = count_coefficients[, "z value"],
  P_Value = count_coefficients[, "Pr(>|z|)"]
)

zero_inflation_table <- data.frame(
  Predictor = rownames(zero_coefficients),
  Estimate = zero_coefficients[, "Estimate"],
  Std_Error = zero_coefficients[, "Std. Error"],
  Z_Value = zero_coefficients[, "z value"],
  P_Value = zero_coefficients[, "Pr(>|z|)"]
)

combined_table <- merge(
  poisson_table, 
  zero_inflation_table, 
  by = "Predictor", 
  suffixes = c("_Poisson", "_ZeroInflation")
)

# Create a new column to identify the model type (Poisson or Zero-Inflation)
poisson_table$Model <- "Poisson"
zero_inflation_table$Model <- "Zero-Inflation"

# Combine the two tables vertically
combined_table <- rbind(poisson_table, zero_inflation_table)

# Create the HTML table using kable
combined_table_html <- combined_table %>%
  kable("html", col.names = c("Predictor", "Estimate", "Std. Error", "Z Value", "P Value", "Model")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Save the table to an HTML file
save_kable(combined_table_html, file = "zip_model_output.html")

#lagged zip
lag_zip_coefficients <- summary(lag_zip_iv_model)$coefficients
lag_count_coefficients <- lag_zip_coefficients$count
lag_zero_coefficients <- lag_zip_coefficients$zero

lag_poisson_table <- data.frame(
  Predictor = rownames(lag_count_coefficients),
  Estimate = lag_count_coefficients[, "Estimate"],
  Std_Error = lag_count_coefficients[, "Std. Error"],
  Z_Value = lag_count_coefficients[, "z value"],
  P_Value = lag_count_coefficients[, "Pr(>|z|)"]
)

lag_zero_inflation_table <- data.frame(
  Predictor = rownames(lag_zero_coefficients),
  Estimate = lag_zero_coefficients[, "Estimate"],
  Std_Error = lag_zero_coefficients[, "Std. Error"],
  Z_Value = lag_zero_coefficients[, "z value"],
  P_Value = lag_zero_coefficients[, "Pr(>|z|)"]
)

lag_combined_table <- merge(
  lag_poisson_table, 
  lag_zero_inflation_table, 
  by = "Predictor", 
  suffixes = c("_Poisson", "_ZeroInflation")
)

# Create a new column to identify the model type (Poisson or Zero-Inflation)
lag_poisson_table$Model <- "Poisson"
lag_zero_inflation_table$Model <- "Zero-Inflation"

# Combine the two tables vertically
lag_combined_table <- rbind(lag_poisson_table, lag_zero_inflation_table)

# Create the HTML table using kable
lag_combined_table_html <- lag_combined_table %>%
  kable("html", col.names = c("Predictor", "Estimate", "Std. Error", "Z Value", "P Value", "Model")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Save the table to an HTML file
save_kable(lag_combined_table_html, file = "lag_zip_model_output.html")

```



# Use tidy() to extract the results from each model (exp zip)
tidy_late_model <- tidy(late_model)
tidy_agg_late_model <- tidy(agg_late_model)
tidy_monthly_lag_iv <- tidy(monthly_lag_iv)
tidy_agg_lag_late_model <- tidy(agg_lag_late_model)

#spearate zip coefficients
zip_coefficients <- summary(zip_iv_model)$coefficients
zip_count_coefficients <- zip_coefficients$count
zip_zero_coefficients <- zip_coefficients$zero

lag_zip_coefficients <- summary(lag_zip_iv_model)$coefficients
lag_zip_count_coefficients <- lag_zip_coefficients$count
lag_zip_zero_coefficients <- lag_zip_coefficients$zero

# Manually create a tidy data frame for the zeroinfl model
zip_tidy <- data.frame(
  term = c(rownames(zip_count_coefficients), rownames(zip_zero_coefficients)),
  estimate = c(zip_count_coefficients[, 1], zip_zero_coefficients[, 1]),
  std.error = c(zip_count_coefficients[, 2], zip_zero_coefficients[, 2]),
  p.value = c(zip_count_coefficients[, 4], zip_zero_coefficients[, 4]),
  model = rep("Zero-Inflated IV", length(c(rownames(zip_count_coefficients), rownames(zip_zero_coefficients))))
)

lag_zip_tidy <- data.frame(
  term = c(rownames(lag_zip_count_coefficients), rownames(lag_zip_zero_coefficients)),
  estimate = c(lag_zip_count_coefficients[, 1], lag_zip_zero_coefficients[, 1]),
  std.error = c(lag_zip_count_coefficients[, 2], lag_zip_zero_coefficients[, 2]),
  p.value = c(lag_zip_count_coefficients[, 4], lag_zip_zero_coefficients[, 4]),
  model = rep("Lagged Zero-Inflated IV", length(c(rownames(lag_zip_count_coefficients), rownames(lag_zip_zero_coefficients))))
)

# Add model labels to the tidy results
tidy_late_model$model <- "Normal IV"
tidy_agg_late_model$model <- "Aggregated IV"
tidy_monthly_lag_iv$model <- "Lagged IV"
tidy_agg_lag_late_model$model <- "Lagged Aggregated"

# Combine the tidy results into a single data frame
combined_tidy_models <- rbind(tidy_late_model, tidy_agg_late_model, zip_tidy, 
                              lag_zip_tidy, tidy_agg_lag_late_model, tidy_monthly_lag_iv)

# Create a table and display it in markdown format
kable(combined_tidy_models, format = "markdown", digits = 3)
